---------------------

author:
- name: Jamaal Green
  affiliation: Portland State University
  email: jamgreen@pdx.edu
- name: Dillon Mahmoudi
  affiliation: University of Maryland, Baltimore County
  email: dillonm@umbc.edu
package: lehdr
date: `r Sys.Date()`
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Getting Started with lehdr}
  %\usepackage[UTF-8]{inputenc}

---------------------

This vignette is a brief introduction to the package includings its installation and making some basic queries. 

# Introduction

**lehdr** is an R package that allows users to draw [Longitudinal and Employer Household Dynamics](https://lehd.ces.census.gov/data/#lodes) Origin-Destination Employment Statistics (LODES) datasets returned as dataframes. The LODES dataset forms the backbone of the US Census's [**OntheMap**](https://onthemap.ces.census.gov/) web app that allows users to track changing spatial employment patterns at a fine geographic scale. While OnTheMap is useful, it is a limited tool that does not easily allow comparisons over time or across geographies. This package exists to make querying the tables that form the OnTheMap easier for urban researchers and practitioners, such as transportation and economic development planners and disaster preparedness professionals.

# Installation

**lehdr** has not yet been submitted to CRAN so installing using **devtools** is required. Additionally, we'll be using **dplyr** and **ggplot2**.

```{r, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman", repos = "http://cran.us.r-project.org"); library(pacman)}
p_load(ggplot2, stringi, stringr, dplyr)
devtools::install_github("jamgreen/lehdr")

```

# Usage
After loading the library, this first example pulls the Oregon (`state = "or"`) 2014 (`year = 2014`), origin-destination (`lodes_type = "od"`), all jobs including private primary, secondary, and Federal (`job_type = "JT01"`), all jobs across ages, earnings, and industry (`segment = "S000"`), aggregated at the Census Tract level rather than the default Census Block (`agg_geo = "tract"`).

```{r usage 1}
library(lehdr)
grab_lodes(state = "or", year = 2014, lodes_type = "od", job_type = "JT01", 
           segment = "S000", state_part = "main", agg_geo = "tract")
```

The package can be used to retrieve multiple states and years at the same time by creating a vector or list. This second example pulls the Oregon AND Rhode Island (`state = c("or", "ri")`) for 2013 and 2014 (`year = c(2013, 2014)` or `year = 2013:2014`).

```{r usage2}           
grab_lodes(state = c("or", "ri"), year = c(2013, 2014), lodes_type = "od", job_type = "JT01", 
           segment = "S000", state_part = "main", agg_geo = "tract")           
```

Not all years are available for each state. To see all options for `lodes_type`, `job_type`, and `segment` and the availability for each state/year, please see the most recent LEHD Technical Document at https://lehd.ces.census.gov/data/lodes/LODES7/.

Other common uses might include retrieving Residential or Work Area Characteristics (`lodes_type = "rac"` or `lodes_type = "wac"` respectively), low income jobs (`segment = "SE01"`) or good producing jobs (`segment = "SI01"`). Other common geographies might include retrieving data at the Census Block level (`agg_geo = "block"`, not necessary as it is default).

# Additional Examples
## Adding at County level signifiers
The following examples loads work area characteristics (wac), then uses the work area geoid `w_geocode` to create a variable that is just the county `w_county_fips`. Similar transformations can be made on residence area characteristics (rac) by using the `h_geocode` variable. Both variables are available in origin-destination (od) datasets and with od, one would need to set a `h_county_fips` and on `w_county_fips`.
```{r example_county_var}
mdrac_jobs <- grab_lodes(state = "md", year = 2015, lodes_type = "wac", job_type = "JT01", segment = "S000")
mdrac_county_jobs <- mdrac_jobs %>% mutate(w_county_fips = str_sub(w_geocode, 1, 5))
```

## Aggregating at County level
To aggregate at theh county level, continuing the above example, we must also drop the original lock geoid `w_geocode`, group by our new variable `w_county_fips` and our existing variables `year` and `createdate`, then aggregate the remaining numeric variables.
```{r example_county_agg}
mdrac_county_jobs <- mdrac_jobs %>% mutate(w_county_fips = str_sub(w_geocode, 1, 5)) %>% 
  select(-"w_geocode") %>%
  group_by(w_county_fips, state, year, createdate) %>% 
  summarise_if(is.numeric, sum)
```

This functionality is also built-in to the package:
```{r example_county_agg2}
mdrac_county_jobs <- grab_lodes(state = "md", year = 2015, lodes_type = "rac", job_type = "JT01", 
           segment = "S000", agg_geo = "county")
```

## Aggregating Origin-Destination 
Aggregating at origin-destination is also built-in, where aggregation takes place on both the `h_geocode` and `w_geocode` variables:
```{r example_county_agg3}
mdrac_county_jobs <- grab_lodes(state = "md", year = 2015, lodes_type = "rac", job_type = "JT01", 
           segment = "S000", agg_geo = "county", state_part = "main")
```

## Functions at Block Group or Tract level
Similarly, built-in functions exist to group at Block Group and Tract levels.
```{r example_county}
mdrac_bg_jobs <- grab_lodes(state = "md", year = 2015, lodes_type = "rac", job_type = "JT01", 
           segment = "S000", agg_geo = "bg")
mdrac_tract_jobs <- grab_lodes(state = "md", year = 2015, lodes_type = "rac", job_type = "JT01", 
           segment = "S000", agg_geo = "tract")
```

